But I can't preregister my research: Improving the reproducibility and transparency of ecology and conservation with adaptive preregistration for model-based research

### 2025-03-03 _analysis/ms.qmd
Update Introduction & Methods

  1. Preregistration is an open-science practice adopted in scientific disciplines that have begun to confront the ‘reproducibility crisis’, which aims to improve research transparency and mitigate questionable research practices within a study. There has been little uptake of preregistration in ecology and conservation, and many modellers and researchers engaged in non-hypothesis testing research both within and outside of ecology have eschewed pre- registration on the grounds that existing templates are irrelevant because they are focused on null-hypothesis significance testing.
  2. We advocate that preregistration can and should be used in model-based research in ecology and conservation, but that preregistration should reflect the norms and practice of the research context in which it is applied, in order to adequately restrict researcher degrees of freedom that are unique to a domain-specific methodology. We took a user-centred design approach to the task of translating preregistration into model- based research contexts for ecology and conservation.
  3. We identified a generalised ‘modelling workflow’ that captures critical analysis decisions in the process of the model development cycle. The workflow informed the content and structure of a preregistration template to be used in model-based research. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the current format and linear process of preregistration. To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we propose a methodology for implementing ‘adaptive preregistration’ leveraging git and GitHub. We tested and evaluated the preregistration template and methodology using a case- study of environmental flows modelling for decision-support in Victoria, Australia.
  4. This research provides a template and methodology for implementing adaptive preregistration of ecological models in either pure or applied settings and will facilitate the wider uptake of preregistration in ecology. Although we focus on ecology and conservation in this paper, the template is extensible to alternative fields and methods, and the proposed methodology for implementing adaptive preregistration can be applied in model- based re- search within science more broadly. Modellers need no longer cry “but, I can’t preregister my research.”
keywords:
  - preregistration
  - transparency
  - reproducibility
  - modelling
  - ecology
  - conservation
  - open-science
  - research software engineering
  - id: aff1
    name: University of Melbourne
  - id: aff2
    name: Arthur Rylah Institute for Environmental Research
format:
  docx: default
  pdf: default
  elsevier-pdf:
    journal:
      name: Journal Name
      formatting: preprint #review or doubleblind
      model: 3p
      layout: onecolumn
      cite-style: authoryear
      # graphical-abstract: "![](abstract.png)"
      # highlights:
        # - Highlight 1
        # - Highlight 2 
        # - Highlight 3
    keep-tex: true
bibliography: references_obsidian.bib

@@

Science's replication crisis, and the subsequent recognition of problems with reliability of scientific research findings, has triggered a 'credibility revolution' spawning a suite of proposed methodological and policy reforms, including *preregistration* [@parker2019]. Preregistration is an open-science practice that aims to distinguish between planned and ad hoc analyses  [@Parker2019] and aims to combat the risk of questionable research practices by requiring researchers to register their methods and analysis plan in a secure and publicly accessible platform prior to collecting and/or analysing data, which cannot be altered after submission ( #todo #citation). The purpose of preregistration is to prevent biased research findings by guarding against questionable research practices (such as selective reporting, p-hacking, HARKing, among others #todo #citation ) by distinguishing what is a genuine *a priori* planned analysis, and what is not, thereby preventing the opportunistic misuse of researcher degrees of freedom, or those many analytic decisions researchers make during the planning, data collection, data analysis and reporting of results in scientific studies (See Box 1 for background). QRPs have been labelled as one of the primary cause of the replication crisis (Shrout and Rodgers, 2018 in [@Fife2022]) and serve to XYZ bias the overall literature, etc ( #todo finish this point).

Despite enthusiastic uptake in some scientific disciplines, such as psychology and medicine ( #todo #citation ), the use of registered reports and preregistration in ecology and conservation more broadly is new and limited [@Desbureaux2021], particularly for model-based research ( #todo Joel Pick preprint?)**. Ecologists have brushed preregistration off on the basis that existing templates are irrelevant since they are focussed on null-hypothesis significance testing**, whereas in applied ecology and conservation, we don't usually explicitly test hypotheses [@Betts2021], but instead use explanatory models, or increasingly predictive models that generate anticipatory predictions, to inform management decisions [@Wood2020]. Model-fitting and evaluation, rather than falsification of hypotheses about natural phenomena, are increasingly prevalent approaches to scientific inquiry in ecology [@Connolly2017]

Preregistration in model-based research has been eschewed on several grounds by modellers and researchers engaged in model-based research both within and outside of ecology. Much of the literature advocating for the use of preregistration assumes a hypothetico-deductive model of scientific inquiry, and the templates used to write preregistrations have been accused of being focussed on null-hypothesis significance testing [e.g. @McDermott2021; @Prosperi2019]. Approaches and advocacy for preregistration of MBR across all scientific disciplines are limited, and focussed have focused on the phases of the modelling process and modelling approaches that are more confirmatory ( #todo see @ioannidis2022). For example, @Cruwell2019b present a template for preregistration of MBR where the model construction and development phase is explicitly and intentionally excluded from the preregistration process, while much of the discussion around preregistration for models has centred on machine learning [@Hildebrandt2018] or predictive modelling [@hofman2023] ( #todo see \[preregistration for modelling\] for more refs).

A second argument against preregistration for model-based research is based on the mischaracterisation of the model development process as a intrinsically exploratory process [@McDermott2021] , and since only the preregistration of confirmatory research, such as hypothesis testing, is considered necessary [@Prosperi2019], there is no need to pre-register model-based research ( #todo #citation). The description of modelling as a wholly exploratory process is a mischaracterisation based on a misunderstanding of sensemaking and reasoning within the modelling process stemming from a lack of clear formal definitions distinguishing confirmatory research from exploratory research [@Flournoy2020; @Devezer2020; @alspaugh2019]. Although modelling includes some tasks and activities that are exploratory in nature, it can be distinguished from wholly *exploratory analysis* in that tends be goal-driven, directed analysis that is less open-ended [@alspaugh2019]. For model-based research, the exploratory/confirmatory divide is an idealised and simplifying dichotomy that, in practice, is more complicated and nuanced and can be considered as a spectrum on which exploratory analysis sits at the analysis is open-ended, high-level and opportunistic end, to goal-driven, low-level and precise analyses at the other [@alspaugh2019]. In practice the divide is often blurred, and researchers will move back and forth between different types of reasoning and inference throughout the modelling process, switching from activity at one end of the spectrum to the other and back again [@alspaugh2019], with deductions leading to abductions as the analysis process proceeds [@alspaugh2019; @Prosperi2019], or even occuring together in a hybridized manner [@Prosperi2019].

When critics of preregistration for MBR describe modelling as intrinsically 'exploratory', what is seemingly being referred to is the iterative, adaptive and non-linear nature of model-development. In practice, modelling proceeds in an adaptive manner by which models are iteratively developed and refined [@Augusiak:2014gz; @Babel2019; @Bennett2013; @Bodner2020]. New analyses are often chosen based on the basis of data exploration and previous analyses of the same data [@Dwork2015], For example, modellers may perform exploratory analyses to validate distributional assumptions are met, such as no over-dispersion or zero-inflation in count-data to ensure valid inference when using particular models [@Campbell2021] ( #todo also see @alspaugh2019 for list of examples). Such practices violate a key tenet of preregistration, that is 'data-independent' analytic decision-making ( #todo #citation). Consequently, the iterative nature of model development, wherein models are continually refined through analysing and reanalysing a dataset and the model, is considered to be in conflict with the practice of preregistration; by requiring researchers to commit to a single static plan that prevents changes or deviations from the original plan it is seen to be forcing adaptive analysis to be *non-adaptive* through 'defining the entire analysis protocol ahead of time' [@Dwork2015; @Benning2019; @MacEachern2019; @McDermott2021]. In short, we can't write off preregistration for modelling on the basis that it is exploratory, because in practice it employs a range of modes of reasoning from induction, deduction, to abduction, with analysts switching between those types of reasoning or even hybridising them.

What *is* important to consider when considering the application of preregistration within MBR is whether the reliability of MBR may be potentially threatened by the potential for Questionable Research Practices (Box 1). Evidence from ecology more broadly has revealed that the rate of QRPs is similar to those in other disciplines [@Fraser:2018cl] and while there has been little study of QRPs in MBR (except @Gould:2020ebg #todo preprint doc & cross-ref here), preliminary evidence reveals that modellers do engage in QRPs such as selective reporting, for example, they may may omit explored analysis paths from their reported analyses if they did not corroborate their conclusions [@Liu2020]. Furthermore, QRPs are likely to arise in conditions where there is incomplete and opaque reporting practices [@Fidler:2018ho], and poor model-reporting practices and an overall lack of transparency about the modelling process are endemic in MBR in ecology. Concerns about the state of model reporting in ecological modelling have been raised repeatedly, arguing that current practices are neither transparent or comprehensive [@Schmolke:2010fd; @deVos:2011uk]. Despite a number of attempts to promote thorough and transparent reporting, (e.g. @Jakeman:2006ii; @Alexandrov:2011iv; @Bodner2020) as a critical component of *Good Modelling Practice*, a culture of adequate research transparency in modelling is yet to be established and remains limited [@deVos:2011uk; @Grimm:2014es]. The lack of transparency in describing ecological models and reporting the modelling process and results provides significant opportunity for undisclosed researcher degrees of freedom, and hence, the potential for Questionable Research Practices.

Although the purpose of preregistration is predominantly conceptualised in terms of distinguishing exploratory from confirmatory research (e.g. @Nosek:2018eg, @munafo2017 and #todo insert other refs), we we adopt a more general stance towards the purpose of preregistration, i.e. that preregistration may mitigate against questionable research practices by reducing the potential for cognitive biases to cause opportunistic misuse of researcher degrees of freedom. In this paper, we advocate that preregistration can and should be used in model-based research in ecology and conservation, but its particular form should reflect the norms and practices of the research context in which it is applied, in order to adequately restrict researcher degrees of freedom that may uniquely emerge within this domain-specific methodology.

The goal of this study is to rethink preregistration to better suit the domain-specific challenges and methodological particularities of model-based research in ecology and conservation. Preregistration cannot simply be ported over to other disciplines and adopted as-is. Different research methodologies, and their domain-specific application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology. If we wish to adopt preregistration for ecological modelling, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology.

We first aimed to build a preregistration template for ecological modelling. We identified a generalised ‘modelling workflow’ that captured critical analysis decisions in the process of the model development cycle. The workflow informed the content and structure of a preregistration template to be used in model-based research. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the current format and linear process of preregistration. To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we devised a methodology for implementing an expanded view of preregistration that we term *Adaptive Preregistration* [sensu @Srivastava2018]. Adaptive preregistration allows for the iterative cycle of model development to be reflected in the preregistration and better accounts for the inherently adaptive nature of model development than current approaches to preregistration in which researchers are committed to a static, unalterable and deterministic view of preregistration. Finally, to assess the feasibility and practicality of preregistration for MBR, we tested and evaluated the preregistration template and methodology using a case-study of environmental flows modelling for decision-support in Victoria, Australia.

------------------------------------------------------------------------

# Box 1: How Researcher Degrees of Freedom become Questionable Research Practices in MBR

Modellers are faced with numerous 'researcher degrees of freedom' [@Wicherts2016] or subjective decisions during the course of modelling that may substantially impact the chosen analysis strategy, and therefore the research findings and conclusions drawn from that analysis [@Hoffmann2021; @Simonsohn2020; @Desbureaux2021; @gouldSameDataDifferent2023]. The set of all plausible or reasonable potential decisions that a researcher could take from the beginning to end of analysis has been described as the 'garden of forking paths' [@Gelman2013]. 'Many analyst'-style studies, whereby multiple separate teams of researchers independently analyse the same dataset and same research question, demonstrate that different pathways in this 'garden of forking paths' can lead to potentially conflicting conclusions [@Young2021; @Liu2020; @gouldSameDataDifferent2023].

Researcher degrees of freedom arise because there is inherent analytic uncertainty in model-based research whether due to incomplete theoretical understanding of underlying ecological processes, or because modelling is an act of simplification and abstraction; there is always some degree of choice about which ecological processes and phenomena should be included in the model, how they should be represented [@Babel2019; @Getz:2017ic]. Further analytic uncertainty may arise due to uncertainty in data processing or choice about which analysis method is best [@Hoffmann2021].

-   *Analytic Uncertainty* [@Walker2003] talks about uncertainty specifically in terms of *model based* decision-making. #todo

When researchers are faced with ambiguous decisions where there is no single decision or strategy acceptable according to scientific standards and consensus, we are likely to select the decision that results in the most 'favourable' finding with "convincing self-justification" [@Simmons:2011iw; @Hoffmann2021]. Researchers routinely execute and explore alternative analysis pathways in service of identifying the most favourable (i.e. publishable) result, selectively reporting only one or several preferred results [@Liu2020; @Simmons:2011iw]. Thus researcher degrees of freedom precipitate 'questionable' research practices, or QRPs, when they remain both undisclosed and exploited opportunistically [@Bakker2018]. Such 'questionable' research practices have been demonstrated to be surprisingly common upon admission among ecologists, with comparable rates to other scientific disciplines [@Fraser:2018cl]. Questionable research practices are categorically distinct from fraudulent practices, arising not from consciously nefarious 'fishing expedition\[s\]' [@Gelman2013], but rather because researchers are motivated actors working within a 'publish or perish' culture who are susceptible to cognitive biases and are willing to change their modelling assumptions when the results conflict with desired outcomes, for example, sincerely and self-convincingly believing that the more favourable model is superior [@Young2021].

QRPs may increase the probability of false positive results, may artificially inflate effect size estimates, and ultimately reduce the replicability of published findings, while encouraging overconfidence in the precision of results and exacerbating the risk of accepting and propagating false facts [@Wicherts2016; @Hoffmann2021; @Roettger2019; @Nissen2016]. Good intentions, and simply being 'aware' of the potential for researcher degrees of freedom to lead to questionable research practices cannot sufficiently mitigate their risk [@Nosek:2018eg].

------------------------------------------------------------------------

@@

We approached the task of translating preregistration into model-based research contexts for ecology and conservation with a user-centred design (UCD) framework. UCD is a philosophy for designing processes, services or products that account for user requirements at all stages of the design process so that the resource generated by this approach is effective and usable for specific people in specific contexts [@Pavelin2012]. Because it is difficult to predict how users will actually engage with a resource [@Pavelin2012] and a designer's intention for preregistration may conflict with the user's purpose and undermine the overarching goal of preregistration [@Pu2019], we heed the advice of Pu et al. [-@Pu2019], who recommend treating preregistration as a UCD problem so that preregistration can effectively meet its intended objectives, despite UCD being typically used within software engineering domains, rather than academic domains.

We borrow from the methodology proposed by Pavelin et al. [-@Pavelin2012] to design, test and refine a preregistration template suitable for ecological modelling, and a methodology for preregistering model-based research (@fig-UCD). The methods description is structured following the key steps illustrated in @fig-UCD.

![Insert caption. Number each step so i can refer to it in the headings and text below. Make it clear in the figure that we're not doing the build interface and conduct usability testing component of implement, nor the evaluate and iterate. we throw that out to the community of practice to do that part.](images/Pasted%20image%2020240314123346.png){#fig-UCD}

## Define Vision & Objectives

> Define the goals of the research under development and identify the audience

The first step in a UCD approach is to articulate the goals, objectives and target audience of the resource, consequently we define the intended purpose of the preregistration template and methodology, the community of users, as well as their research contexts and use-cases. Finally, we explicate the aspirations and desirable attributes of the template and methodology. Much like in 'Good Modelling Practice', articulating the overarching purpose, intended audience, as well as the objectives and aspirations will guide the development and evaluation of preregistration for model-based research in ecology and conservation, with important implications for the design of the template -- from its structure, content, user-interface and broader workflow in which it is designed to be used.

### Articulating Intended purpose of preregistration template & methodology

We define the purposes of the template as:

1\. *To delimit researcher degrees of freedom* \> when conducting model-based research in applied ecology and conservation so as to mitigate the risk of questionable research practices in order to improve the reliability and credibility of this research.

2\. *Increase research transparency* \> of model-based research in ecology and conservation. *Research transparency* refers to both production transparency — which includes research artefacts like open-access data and materials, or data collection procedures — and analytic transparency — which describes a complete account of how the analytic conclusions are drawn from the data [@Lupia2014].

### Target user community and their research-contexts

We aim for the template and preregistration process developed in this study to be used for model-based research, which we define as any research, empirical or applied, that uses quantitative modelling to answer its research question [@Brudvig2017], where *quantitative modelling* may be defined as any modelling that "describes and/or forecast(s) the behaviour of a system using mathematical and statistical concepts, and whose parameters and their relationships are expressed as quantities" [@Garcia-Diaz2019].

We define model-based research loosely in order to capture the breadth of different modelling goals (e.g. exploration, inference and prediction; @Tredennick2021) and type of research contributions (e.g. development of a new method, etc #todo see @Moallemi2019 and across both pure and applied ecological modelling.

### Specifying Preregistration use-cases

Pu et al. [-@Pu2019] describe two key use-cases of preregistration, depending on the role assumed in interacting with the preregistration. Users have different requirements of preregistration depending on the role that they assume, that is, study authors and modellers have a different use-case than reviewers, editors and readers, however this latter use-case is often not explicitly designed and accounted for in current platforms/infrastructure/processes [@Pu2019]. Reviewers and editors are usually tasked with the process of 'checking' the preregistration, comparing the final manuscript to its preregistration and verifying that the preregistered analyses are conducted and reported as specified [@Pu2019]. However existing formats are designed primarily for the use-case of the author - they are text-based static documents that are ill-suited to checking the reported study against the preregistration.

Although preregistration authorship is the primary target use-case of the preregistration template and methodology developed in this study, we aim to consider and incorporate features into the preregistration process, such as template formatting, that will facilitate preregistration checking to also address non-authorship use-cases.

### Articulating desirable attributes of the preregistration template and adaptive preregistration methodology

The preregistration template should capture critical decision-points that may influence the overall model and analysis as well as any recommendations or decisions drawn from analysis of the model, thus delimiting researcher degrees of freedom at these junctures in the modelling decision process. The structure and content of the template is necessarily opinionated, encouraging Good Modelling Practices, while reflecting community norms and practices in model-development while still meeting the objective of mitigating the risk of Questionable Research Practices by delimiting researcher degrees of freedom. The template needs to be generalisable enough to apply to a range of common modelling purposes, methods and research contributions, and flexible enough that if a particular decision is not applicable for the planned study, that those items may be removed or altered, and other items may be added onto the template.

Secondary to the aim of mitigating QRPs, is that the the template should also enhance the research transparency of the preregistered study, ensuring that analysis choices are documented with sufficient transparency to facilitate the reader of the preregistered completed study in making informed assessments about the credibility and reliability of the model and any associated analysis and recommended interventions. Borrowing from @Bodner2020's framework to increase the accuracy, reliability and transparency of predictive models, in order to increase transparency of model-reporting, the template should explicitly outline critical model choices and their assumptions, steps in the modelling process, and explicit expectations about the outputs and analysis of the model.

**Adaptive Preregistration Methodology**

The Adaptive Preregistration Methodology aims to capture the entire modelling process, including model development and construction, rather than beginning at the point of analysing a pre-existing and parameterised model for deductive inference / prediction. Critically, the methodology must not preclude the iterative and adaptive nature of model construction, while simultaneously restricting the types of data-dependent analytic decision-making that may compromise the reliability and validity of the model and its contingent analysis and conclusions.

## User Research & Analysis

> Characterise users and their needs

On the 4th and 5th of March 2020, we facilitated a series of structured activities at the Arthur Rylah Institute for Environmental Research which aimed to inform the development of two preregistration templates, one for field-work and sampling design, the other for modelling and data-analysis. We split the activities into two groups, to better focus discussion for each of these templates.

-   [ ] <mark style="background: #FFF3A3A6;">Note: might need to reframe this and omit the field study stuff, but should probably state that we did this but decided not to pursue?</mark>

The objectives of the collaborative workshop activities were to aid in: 1. Understanding how modelling is done in practice to inform and tailor the preregistration templates to adequately capture key decision-points in the scientific modelling process; 2. Understanding how science in practice -- whether due to logistical, technical, feasibility or normative constraints -- differs from some idealised practice, in order to: 3. Identifying challenges and roadblocks to the use and uptake of preregistration in among ecological modellers, both in empirical and applied research contexts; and to 4. Using the above knowledge to inform the preregistration template and process design so they meet their intended purpose, and are adopted by our target user community for their particular use-cases.

### Workshop 1: Identifying Ecological Modelling Workflows, Norms, and Practice

>>>>

  1.	Preregistration is an open-science practice which aims to improve research transparency and mitigate questionable research practices, like cherry picking results. It helps protects against cognitive biases, like hindsight bias, that can influence how study outcomes are interpreted. There has been little uptake of preregistration in ecology and conservation, arguably because existing pre-registration templates focus on null-hypothesis significance testing whereas ecology and conservation often rely on different types of statistical modelling.
  2.	We argue that preregistration in model-based research in ecology and conservation is both possible and beneficial, using templates adapted for domain-specific methodologies. We applied a user-centred design approach to translate the concept of preregistration into model-based research practice for ecology and conservation.
  3.	We developed a generalised ‘modelling workflow’ to capture critical analysis decisions in the model development cycle, informing a preregistration template for model-based research. The iterative nature of model development is incompatible with existing, linear preregistration templates, creating a barrier to implementing preregistration for model-based research.
  4.	To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we propose, test and evaluate a methodology for ‘adaptive preregistration’, using a case-study of environmental flows modelling for decision-support in Victoria, Australia.
  5. This research provides a template and methodology for implementing adaptive preregistration of ecological models. Although we focus on ecology and conservation in this paper, the concept of adaptive preregistration, and the templates developed here, could be applied in model-based research in other scientific disciplines within science more broadly. Modellers need no longer cry “but I can’t preregister my research.”

keywords:
  - preregistration
  - transparency
  - reproducibility
  - modelling
  - ecology
  - conservation
  - open-science
  - research software engineering
affiliations:
  - id: aff1
    name: University of Melbourne
  - id: aff2
    name: Arthur Rylah Institute for Environmental Research
format:
  # preprint-typst:
    # fontsize: 11pt
    # mainfont: Libertinus Serif 
  # docx: default
  preprint-docx:
    toc: false
    number-sections: true
    highlight-style: github
  elsevier-pdf:
    journal:
      name: Journal Name
      formatting: preprint #review or doubleblind
      model: 3p
      layout: onecolumn
      cite-style: authoryear
      # graphical-abstract: "![](abstract.png)"
      # highlights:
        # - Highlight 1
        # - Highlight 2 
        # - Highlight 3
    keep-tex: true
bibliography: "references_obsidian.bib"
csl: methods-in-ecology-and-evolution.csl
---

@@

Over the past ten years or so, a ‘replication crisis’ in some scientific fields has exposed problems with the reliability of research findings [@Neoh2023; @OpenScienceCollaboration:2015cn; @camererEvaluatingReplicabilitySocial2018; @Ioannidis:2014cm]. One such problem is the prevalence of Questionable Research Practices (QRPs), including p-hacking, Hypothesising After Results are Known (HARKing), and cherry-picking [@Wicherts2016]. QRPs increase the probability of false positive results, artificially inflate effect size estimates, and ultimately reduce the replicability of published findings, while encouraging overconfidence in the precision of results and exacerbating the risk of accepting and propagating false facts [@Hoffmann2021; @Roettger2019; @Nissen2016]. Self-reports from researchers show that QRPs are surprisingly common across disciplines [@John:2012eo; @Agnoli:2017kl; @Makel2023], including in ecology [@Fraser:2018cl; @kimmel2023]. Although these self-report surveys of QRPs are heavily focused on statistical significance testing research employing p-values, some broader concepts, like selective reporting/cherry picking, have counterparts in other inferential frameworks.

QRPs are precipitated when researchers opportunistically or systematically misuse ‘researcher degrees of freedom’ without disclosure [@Bakker2018; @Wicherts2016]. Researcher degrees of freedom refer to the many decisions and alternative analytic choices researchers make throughout the research process [@Simmons:2011iw; @Wicherts2016]. Analytic decisions are typically subjective but methodologically defensible, however, choices and omissions taken at each ‘fork’ may collectively and significantly influence the final ‘analysis strategy’ [*sensu* @Hoffmann2021], ‘specification’ [*sensu* @Simonsohn2020], or ‘modelling path’ [*sensu* @hamalainen2016]. The set of all (plausible or reasonable) possible decisions that a researcher could make over the entire analysis from beginning to end has been termed collectively as the ‘garden of forking paths’ [@Gelman2013] or the ‘multiverse’ [@Steegen2016]. ‘Many-studies’ have now demonstrated the consequences of the ‘garden of forking paths’, whereby alternative analytical decisions can lead researchers to make vastly different interpretations of results, with conflicting conclusions, even when investigating the same research question using the same underlying dataset [@Gould2025; @Hoogeveen2022; @Silberzahn2018; @Desbureaux2021; @Young2021].

Good intentions and simply being ‘aware’ of the potential for researcher degrees of freedom to lead to QRPs cannot sufficiently mitigate their risk because of the unconscious nature of their origin [@Nosek:2018eg; @Zvereva2021]. Fortunately, there are some potential solutions to this problem. Preregistration is an open-science practice that aims to distinguish between planned and ad hoc analyses [@Parker2019] thereby restricting opportunities for trialling analytical alternatives (‘researcher degrees of freedom’) and making any deviations from the preregistered plan transparent [@Wicherts2016]. Preregistration requires researchers to register their methods and analysis plan in a secure and publicly accessible platform prior to collecting and/or analysing data, which cannot be altered after submission [@Parker2019]. Preliminary empirical evidence indicates that studies with preregistered analysis plans are less likely to have positive or significant results and are more transparently reported [@brodeur2024; @toth2021], especially when the preregistration template includes detailed instructions and requirements [@Bakker2018] or the preregistered study is published as a ‘Registered Report’ [@scheel2021] – a new publication format that aims to combat selective reporting and publication bias wherein the peer-review process and decision to publish is made prior to the results are known.

One potential objection to preregistration for modelling is that model development is an intrinsically exploratory process [@McDermott2021], while preregistration is for confirmatory research [i.e. hypothesis or theory testing, @Prosperi2019]. These conceptions of modelling and preregistration are not entirely wrong, but they are oversimplified. In practice, the distinction between exploratory and confirmatory activities is not a sharp one [@Flournoy2020; @Devezer2020; @alspaugh2019]. Modelling includes some tasks and activities that are exploratory in nature, but it can be distinguished from wholly *exploratory analysis* in that modelling is often goal-driven and directed [@alspaugh2019]. In practice, researchers move back and forth between different types of reasoning and inference throughout the modelling process, with deductions leading to abductions as the analysis proceeds [@alspaugh2019; @Prosperi2019] or even occurring together in a hybridized manner [@Prosperi2019]. This bears a strong resemblance to more general philosophical observations about the iterative nature of science [@Chang2004].

In this iterative and non-linear process of modelling, initial analyses often inform future analyses of the same data whereby the modeller’s understanding of the problem and the model itself are incrementally adjusted in light of intermediate results from previous decision points [@hamalainen2016; @Schweinsberg2021; @Dwork2015]. Modellers are typically uncertain about which model is most appropriate for the data [@popovic2024] and are reluctant to risk registering a modelling strategy that may be incompatible with the data [@Roettger2019]. For example, modellers may perform exploratory analyses to validate distributional assumptions are met [@Campbell2021], which may result in subsequent changes to the model specification. Should data-dependent decisions like these be interpreted as violating a key tenet of preregistration, that is, failing to define the entire analysis protocol ahead of time [@Dwork2015; @Benning2019; @MacEachern2019]? Our answer to this is ‘no’, and within this paper we outline a methodology for implementing an expanded view of preregistration called *Adaptive Preregistration* [*sensu* @Srivastava2018], which captures decision points and reasoning about modelling choices, but which embraces the data-dependency practices that constitute the modelling development process. This goes beyond the work of others who offer discussion and guidance about when and how to deviate from a preregistration [@Lakens2024; @willroth2024], instead providing an approach to preregistration embodying the principles of *registered flexibility* [@Roettger2019; @Benning2019; @Srivastava2018] and iterative or *interim preregistrations* [@ioannidis2022; @hofman2023].

In this study we aim to translate preregistration for application to model-based research, specifically aimed at ecology, conservation and related disciplines. Here, we first identify a generalised ‘modelling workflow’ that captured critical analysis decisions in the model development cycle and use it to build an Adaptive Preregistration template that can work for ecological modelling. We then tested and evaluated our template using a case-study of environmental flows modelling for decision-support in Victoria, Australia. This is the first attempt at developing a fully-fledged preregistration template designed for model development and evaluation that we are aware of, and we propose it can be extended to work in disciplines outside of ecology.

@@

We modified a typical preregistration into model-based research contexts with a user-centred design framework, where users were engaged at all stages to produce an implementation of preregistration that is fit for purpose [@Pavelin2012, Fig. 1].

<!--# Insert Figure 1 Here -->

Figure 1 Process for translating preregistration to ecological modelling research [adapted from @Pavelin2012]. Note that the ‘Evaluate and Iterate’ component of the process was not undertaken in this study

## Problem Formulation: Defining Vision & Objectives for a preregistration template for model-based research

We first specified the research context, purpose, target users and use-cases, per (Pu et al. 2019a; Pavelin et al. 2012). Our context is model-based research, defined broadly as any research that uses quantitative modelling to answer its research question (Brudvig 2017). The purposes of the preregistration template and methodology are to:

1.  **Delimit 'researcher degrees of freedom'** [@pu2019a], to mitigate the the risk of questionable research practices when conducting model-based research in applied ecology and conservation.
2.  **Increase ‘research transparency’ of ecological modelling,** including both:
    1.  *production transparency* — including research artefacts like open-access data and materials, or data collection procedures [@Lupia2014],
    2.  *analytic transparency* — a complete account of how analytic conclusions are drawn from the data [@Lupia2014] where model choices, steps in the modelling process, assumptions and expectations about model outputs are clearly articulated [@Bodner2020].

We specified two use-cases for our Adaptive Preregistration template:

1.  Study authors or modellers seeking to preregister their modelling study,
2.  Editors, reviewers, and/or readers of the completed preregistration [@Pu2019; @pu2019a], who are tasked with verifying that analyses and findings reported in the final manuscript are consistent with the preregistered analysis plan.

Existing preregistration formats are designed primarily for study authors; they are static text-based documents that do not easily facilitate ‘preregistration checking.’ We designed our template to address this.

## Designing the General Modelling Workflow

We conducted a literature review to identify a modelling workflow suitable for application to a range of different modelling goals, e.g. exploration, inference and prediction, including both explanatory and anticipatory predictions [@Tredennick2021; @Mouquet2015; @Maris2018; @Shmueli:2010ec]; research contributions, e.g. development of new methods, or application / extension of existing methods and model transfers [@Yates2018]; types of models, e.g. phenomenological vs. mechanistic [@Connolly2017; @White2019; @lewis2023], across basic and applied contexts and across fields / disciplines in ecology and conservation [@Carmel2013; @Brudvig2017] while offering prescriptive guidelines based on ‘Good Modelling Practice’ [@Augusiak:2014gz]. The final output was a workflow reflecting a balance between idealised and actual modelling in practice.

We developed an Adaptive Preregistration methodology, so that users have the flexibility to adapt the preregistration template to their own needs and processes [see user guide <https://egouldo.github.io/EcoConsPreReg/>, @gould2024]. This can be facilitated in two ways: 1). The preregistration author can supply analytical or modelling heuristics or strategies that depend on the outcome of the previous decision-point or analysis, for example, decision-trees for certain points in the analysis or items in the preregistration template could be preregistered prior to analysing [@Baldwin2022a]. 2) using interim preregistrations that mark the transition between phases of modelling and analysis as different parts of the data are observed [@Srivastava2018].

Modellers can preregister ‘plans to deploy flexible strategies’ [@Srivastava2018] describing heuristics consisting of multiple analysis or modelling strategies, whose execution depends on the outcome of previous decision-points or analyses. For example, the modeller can preregister a ‘decision-tree’ [@Baldwin2022a] consisting of pre-defined rules about when a particular modelling strategy or decision should be implemented. Modellers can specify a ‘flexible’ strategy within their preregistration when faced with methodological or analytic uncertainty that cannot be resolved without observing parts of the data or by engaging in preliminary analyses, model checking, or other data-dependent decisions.

The modeller may follow an iterative process of preregistration in parallel with modelling, consisting of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observe [@Srivastava2018]. As they proceed through the model development process the researcher shifts from ideation, preregistration and execution of the preregistered analysis plan and back again, generating interim preregistrations at different points within the model development process, depending on observed outcomes of the flexible heuristics described in the interim preregistration. The analyses executed in practice will depend on the observed outcomes of the registered flexible analyses.

Last, to facilitate transparency and preregistration checking of flexible modelling and analysis strategies, the results of any final and preliminary analyses must be linked back to the preregistered flexible strategy such that the link between a particular triggered analysis decision, and the analysis outcome that triggered the decision is explicit. Fig. 2 demonstrates a generalised process for using git and GitHub as the vehicle for transparently documenting Adaptive Preregistration of a modelling study. The procedure leverages [GitHub’s tag and release feature](https://help.github.com/en/github/administering-a-repository/releasing-projects-on-github) in conjunction with [Semantic Versioning](https://egouldo.github.io/EcoConsPreReg/(https:/semver.org/)) to track, document and collaborate on changes to the preregistration document. The preregistration is stored alongside modelling code and results with GitHub issues [see @braga2023 for explanation] tracking analysis tasks and results for a discrete task described in the preregistration and providing the opportunity to document interpretation of discussion informing the rationale for potential changes or updates to the preregistration that generate interim preregistrations. The use GitHub issues is particularly handy for registered flexible analyses: the outcome of the registered flexible analysis is recorded within that issue, linked to the described heuristic, and modellers can use the discussion features to document interpretation of those results, and provide interpretation to explain which registered decision alternative is to be selected in light of those results.

## Designing the Preregistration Template

We facilitated a workshop with 10 ecologists and ecological modellers. A key task was to identifycommon modelling workflows and critical decision-steps in model development. ‘Critical’ decision-steps are points in the modelling process where researchers, or modellers, make analytical decisions that could change the output of the analysis. In this activity, workshop participants first individually reflected on their own scientific process for a recent or memorable research project, listing the steps undertaken throughout the process. Next, participants collaboratively mapped the decision steps in their own personal modelling workflows onto a pre-prepared modelling workflow template (<https://osf.io/fgd23>), informed by our general workflow. We then collaboratively reviewed and refined the modelling workflow into a standard set of key modelling phases and commonly implemented decision steps.

<!--# Insert Figure 2 - Adaptive Preregistration Workflow -->

Figure 2 - Implementing Adaptive Preregistration. The preregistration (blue steps) and modelling and analysis processes (orange steps) operate in parallel. Purple tags represent git tags, corresponding to incremental versions of the preregistration.  Any changes to the analysis plan receive a GitHub release and major version number increment. Step 3 corresponds to the initial preregistration document, Step 6 corresponds to an interim preregistration, and Step 8 corresponds to the final preregistration. The execution of the Final Preregistration occurs in Step 9.
